% -------------------------------------------------------
%  English Abstract
% -------------------------------------------------------

\begin{latin}

\begin{center}
\textbf{Abstract}
\end{center}
\baselineskip=.8\baselineskip

One of the main challenges in robust Persian automatic speech recognition is the mismatch between conventional training data and real speech from telephone calls and Voice over IP (VoIP) systems. This thesis presents and evaluates a reproducible approach for adapting Whisper-small to these conditions. For this purpose, a collection of paired, aligned clean and degraded samples is created from Persian speech. Using recorded random seeds, the degradation process simulates conditions such as environmental noise, gain variation, bandwidth restriction, real compression with audio codecs through FFmpeg, and packet loss. Whisper-small is then fine-tuned once on ordinary data and once on a mixture of ordinary and degraded data.

A dual-view model is also investigated. For each degraded input, the model produces two log-Mel representations: one directly from the original signal and the other from its enhanced version. The speech enhancement component consists of a residual U-Net with temporal modeling. After both representations are processed by a shared Whisper encoder, their information is combined using bidirectional cross-attention and a time--feature gate. Training proceeds in three stages: the enhancement component is first prepared independently, the remaining components are then trained, and finally the entire model is jointly optimized end to end.

The models are evaluated on five Persian datasets, including real telephone and VoIP speech from AGFarsdat. The results show that multi-condition training delivers the best overall performance, reducing the aggregate word error rate from 23.27\% to 17.83\%. The fusion model also outperforms ordinary Whisper, with a word error rate of 18.44\%, but does not surpass the multi-condition model. Its character error rate is 14.04\%, which is higher than those of both single-stream models. A mean gate weight of 0.9961 for the original representation shows that the gate relies almost entirely on the original stream, leaving the enhanced stream with a very limited role. The fusion model achieves the lowest word error rate on AGFarsdat; however, without statistical testing, its 0.42 percentage-point advantage over the multi-condition model cannot be considered conclusive or stable. Overall, the results show that, in the current setting, traceable telecom-oriented data augmentation combined with multi-condition training is more effective than the more complex enhancement-and-fusion approach.

\bigskip\noindent\textbf{Keywords}:
Persian automatic speech recognition, telephone speech, Voice over IP, Whisper, multi-condition training, speech enhancement, dual-view fusion

\end{latin}
